### 2022-08-27 main/chapter_searches.tex
add some references for stats section; tweak some data analysis; improve chapter name

\begin{singlespacing}
\chapter{Searches in ATLAS}
\label{chapter:searches}
%
\begin{epigraphs}
\qitem{%
One way is to make it so simple that there are \emph{obviously} no deficiencies
and the other way is to make it so complicated that there are no \emph{obvious}
deficiencies.%
}%
{Tony~Hoare,
\textit{The Emperor’s Old Clothes},
1981~\cite{hoare2007emperor}}
\end{epigraphs}
\end{singlespacing}

@@

How Bayes' rule \emph{should} be used in scientific data analysis is to extract
useful information from the data.
If the likelihood function itself is interpretable, then it should be reported
directly.
When it is not, perhaps due to a complicated parametric modelling, then one
should abstract those parameters away by reporting evidence values for
predictive models, which must of course be clearly stated and interpretable if
they are to be useful.
Data act only through the likelihood function, and reporting a posterior
can act to obfuscate those data.

Posterior probabilities are, however, useful when digesting data to make
inferences about the world.
Reporting likelihoods and evidence values allows readers to assign priors as
they wish.
For model comparison between background $h=B$ and signal $h=S$, this is clearly
expressed in the ratio form
% likelihood ratio
\begin{equation}
\hphantom{\quad \quad \mid\mid \mathrm{context}}
\frac{\prob{S}{d}}{\prob{B}{d}}
=
\frac{\prob{d}{S}}{\prob{d}{B}}
\times
\frac{\probc{S}}{\probc{B}}
\quad \quad \mid\mid \mathrm{context}
,
\end{equation}
where we use the double bar, $\mid\mid$, to refer to any contextual information
a reader might wish to include.
Whenever the evidence ratio is independent of that context, it factors out from
the prior (odds) ratio.
In this sense, likelihood (or evidence) ratios are scale factors that update
any odds ratio by the same multiplication.

@@

Phrased succinctly by Skilling~\cite{skilling2008rant},
``we SHOULD use Bayes by assigning subjective priors and inspecting evidence
values.''
This framing of probabilistic data analysis agrees exactly with our standard
practice of histogramming simulations,
as illustrated in Figure~\ref{fig:searches_sig_bkg_prior_likelihood}.
Signal and background models define priors, and their (normalized) bin yields
are the evidence values that we use to compare the models.
Finally, corresponding posteriors are the normalized shapes those samples take
within the observed histogram bin, with which we do not compare the models,
but which could in principle be used for secondary interpretations.

@@

In decision theory, we want a mathematical procedure that identifies the
optimal decision to make in a given context.
Although this theory has a long history, we point specifically to the result
by Wald~\cite{
wald1947bayes,
wald1950bayes,
jaynes2003probability
}
that the optimal solution to any decision problem can be framed as the
minimization of an expected cost function, or risk:
\begin{equation}
\label{eqn:searches_bayes_decision_rule}
R(t) = \int\! c(t, x) \,\mathrm{d}p(x)
,
\end{equation}
where we mean `expected' in the mathematical mean sense of a `mean' average.
This cost function $c(t, x)$ evaluates the badness of outcomes for a given
decision $t$ and states of the world $x$ to which context must assign prior
probabilities $p(x)$.
For a given cost function and prior, one should choose the decision $t$ that
minimizes $R(t)$.
More important than this Wald's mathematical proof is that this coincides with
intuition --- when considering decisions, it feels normal to imagine what might
occur in each case and consider how good or bad the plausble outcomes are.

@@

This theory does not, however, define how to construct that cost function.
Like the prior, it must come from context.
Indeed, the two are intertwined --- decision-making machines and organisms
are capable of learning to make good decisions without constructing either
explicitly.
Towards our goal of usefully reporting data, such decision theory has been
applied to argue for ways to quote properties of posterior distributions.
For example, a minimizing the expected absolute error $|x - \check{x}$ leads to
reporting the posterior median, and minimizing the expected square error
$(x - \check{x})^2$ leads to reporting the posterior mean.
Had the rules of our bean counting game rewarded you based on such distances,
these results could have been good guesses. But those weren't the rules.

When arguing that we should report likelihoods and evidence values for our
data, we have not explicitly constructed cost functions, nor have we made
quantitative predictions of future events.
But we can qualitatively consider them in this decision theory framework ---
we imagine that future scientists will want to reuse our data, and would be
pleased (attain a small cost) if they were able to reinterpret the results for
their own models.
We can therefore reason that likelihoods would have more utility than
alternatives such as (point) posterior estimates.

% connection to frequentist section
An amusing feature of Wald's theory is that it was formulated from a
perspective on probability in which,
``In most applications, however, not even the existence of an a priori
distribution can be postulated''~\cite{wald1950bayes}.
That might seem problematic, since the risk is the expectation value of a cost
function over a prior (or \emph{a priori}) distribution.
Wald was nonetheless interested in these decision rules for their
``completeness'' that he had proven, and named them ``Bayes'' decision rules
despite approaching the theory from a stalwartly non-Bayesian perspective.

@@

\TODO{
Fisher:
criteria for estimators,
(consistency, efficiency, sufficiency)
https://royalsocietypublishing.org/doi/pdf/10.1098/rsta.1922.0009
testing, p-value,
likelihood, also ratios of p-values (here named probable errors)
https://digital.library.adelaide.edu.au/dspace/handle/2440/15169
tail area = significance
http://psychclassics.yorku.ca/Fisher/Methods/
}

\TODO{
Fisher:
``We can know nothing of the probability of hypotheses or of hypothetical
quantities.''
}

>>>>

\begin{singlespacing}
\chapter{Data analysis and \atlas\ searches}
\label{chapter:searches}
\begin{epigraphs}
\qitem{%
One way is to make it so simple that there are \emph{obviously} no deficiencies
and the other way is to make it so complicated that there are no \emph{obvious}
deficiencies.%
}%
{Tony~Hoare,
\textit{The Emperor’s Old Clothes},
1981~\cite{hoare2007emperor}}
\end{epigraphs}
\end{singlespacing}

@@

How Bayes' rule \emph{should} be used in scientific data analysis is to extract
useful information from the data.
If the likelihood function itself is interpretable, then it should be reported
directly.
When it is not, perhaps due to a complicated parametric modelling, then one
should abstract those parameters away to summarize the likelihood.
Properties such as local maxima or regions in which it exceeds meaningful
thresholds can be useful.
Evidence values for predictive models can also be useful if those models are
clearly stated and relevant to readers' interests.
Data act only through the likelihood function, and reporting a posterior
can act to obfuscate those data.

Posterior probabilities are, however, useful when digesting data to make
inferences about the world.
Reporting likelihood functions and evidence values allows readers to assign
priors as they wish.
For model comparison, for example between background $h=B$ and signal $h=S$
models, the analysis is conveniently linearized by expressing it in $\log$
ratio form:
\begin{equation}
\hphantom{\quad \quad \mid\mid \mathrm{context}}
\phi(d) =
\log\frac{\prob{S}{d}}{\prob{B}{d}}
=
\log\frac{\prob{d}{S}}{\prob{d}{B}}
+
\log\frac{\probc{S}}{\probc{B}}
\quad \quad \mid\mid \mathrm{context}
,
\end{equation}
where we use the double bar, $\mid\mid$, to refer to any contextual information
a reader might wish to include.
When the evidence ratio is independent of that context, as it usually is,
it factors out from the prior (odds) ratio.
In this sense, likelihood (or evidence) ratios are scale factors that update
any odds prior ratio by the same multiplication, and that on this $\log$ scale
that means just adding a constant.

From such $\log$ ratios one individual probabilities are recovered through the
logistic function:
\begin{equation}
\label{eqn:searches_logistic}
\prob{S}{d} = \frac{1}{1 + e^{-\phi(d)}}
.
\end{equation}
Physically, this is analogous to the Boltzmann distribution where the energy
of each model is its negative $\log$ probability, with unit inverse
temperature~\cite{
pmlr-v2-ranzato07a,
skilling2017david
}.
With multiple models in play, it is also called the `softmax'
function~\cite{MurphyKevinP.2012Mlap}.

@@

Phrased succinctly by Skilling~\cite{skilling2008rant},
``we SHOULD use Bayes by assigning subjective priors and inspecting evidence
values.''
This framing of probabilistic data analysis agrees exactly with our standard
practice of histogramming simulations,
as illustrated in Figure~\ref{fig:searches_sig_bkg_prior_likelihood}:
signal and background models define priors, and their (normalized) bin yields
are the evidence values that we use to compare the models.
Finally, their corresponding posteriors are the normalized shapes those samples
take within the observed histogram bin, with which we do not compare the
models, but which could in principle be used for secondary interpretations.

@@

When considering what decision to make, it feels natural to consider what
might result from each possible action and to judge that action on how good or
bad its results would be.
Mathematical decision theory agrees with this intuition---
a famous theorem by Wald~\cite{
wald1947bayes,
wald1950bayes,
jaynes2003probability
}
shows that the optimal solution to any decision problem can be always be
formulated as the minimization of an expected cost function, or risk:
\begin{equation}
\label{eqn:searches_bayes_decision_rule}
R(t) = \int\! c(t, x) \,\mathrm{d}p(x)
,
\end{equation}
where we mean `expected' in the mathematical mean sense of the `mean' average.
This cost function $c(t, x)$ evaluates the badness of outcomes for a given
decision $t$ and states of the world $x$ to which context must assign prior
probabilities $p(x)$.
For a given cost function and prior, one should choose the decision $t$ that
minimizes $R(t)$.

@@

This theory does not, however, define how to construct that cost function.
Like the prior, it must come from context, and indeed the two are intertwined
in the risk;
decision-making machines and organisms are capable of learning to make good
decisions without constructing either cost or prior explicitly.
Towards our goal of usefully reporting data, such decision theory has been
applied to argue for ways to quote properties of posterior distributions.
As examples, minimizing the expected absolute error $|x - \check{x}|$ leads
to reporting the median, and minimizing the expected square error
$(x - \check{x})^2$ leads to reporting the mean.
Had the rules of our bean counting game rewarded you based on such distances,
these results could have been good guesses. But those weren't the rules.

When arguing that we should report likelihoods and evidence values for our
data, we have not explicitly constructed cost functions, nor have we made
quantitative predictions of future events.
But we can qualitatively consider them in this decision theory framework ---
we imagine that future scientists will want to reuse our data, and would be
pleased (attain a small cost) if they were able to reinterpret the results for
their own models.
We can therefore reason that likelihoods would usually have more utility than
alternatives such as point posterior averages.

% connection to frequentist section
An amusing feature of Wald's theory is that it was formulated from a
perspective on probability in which,
``In most applications, however, not even the existence of an a priori
distribution can be postulated''~\cite{wald1950bayes}.
That might seem problematic, since the risk is the expectation value of a cost
function over a prior (or \emph{a priori}) distribution.
Wald was nonetheless interested in these decision rules for their
``completeness'' that he had proven, and named them ``Bayes'' decision rules
despite approaching the theory from a stalwart non-Bayesian perspective.

@@

\TODO{
Lehmann~\cite{lehmann2005testing};
p. 26 poisson inadmissable example;
L-unbiased with given loss function;
}

\TODO{
Fisher:
criteria for estimators,
(consistency, efficiency, sufficiency)
~\cite{fisher1922estimators}
testing, p-value,
likelihood, also ratios of p-values (here named probable errors)
(introduced the term likelihood? cite?)
~\cite{fisher1921probable}
tail area = significance
~\cite{fisher1925smrw}
}

\TODO{
Fisher:
``We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations''
~\cite{fisher1921probable}
Fisher:
``absurdly academic''
~\cite{fisher1956statistical}
% \begin{epigraphs}
% \qitem{%
% We can know nothing of the probability of hypotheses or hypothetical
% quantities.
% On the other hand we may ascertain the likelihood of hypotheses and
% hypothetical quantities by calculation from observations%
% }%
% {Ronald A. Fisher, 1921~\cite{fisher1921probable}}
% \end{epigraphs}
There is something horrifying in the ideological movement represented by the
doctrine that reasoning, properly speaking, cannot be applied to empirical data
to lead to inferences valid in the real world.
~\cite{fisher1956statistical}
}

\begin{singlespacing}
\begin{displayquote}
We can know nothing of the probability of hypotheses or hypothetical
quantities.
On the other hand we may ascertain the likelihood of hypotheses
and hypothetical quantities by calculation from observations
\end{displayquote}
\end{singlespacing}
~\cite{fisher1921probable}

\begin{singlespacing}
\begin{displayquote}
There is something horrifying in the ideological movement represented by the
doctrine that reasoning, properly speaking, cannot be applied to empirical data
to lead to inferences valid in the real world.
\end{displayquote}
\end{singlespacing}
~\cite{fisher1956statistical}

\TODO{
Nayman-Pearson:
Seek good test statistics --
likelihood ratio intuitive for two-hypothesis comparison
prove it optimal in a reasonable sense
idea of long-run mostly not wrong
recede to maximum likelihood ratios
~\cite{neymanpearson1928max}
``for which hypothesis A is more probable and those for which it is less
probable (sic!)''
unbiased (unbiassed) test differs from unbiasedness of estimator
``Then the two corresponding densities at the sample point
are (Al) 6-42 x 10-20, (A2) 5 18*x 1O-4 and the ratio of the likelihood of Hypothesis
Al to- that of A2 is 1P24 x 10-16. This ratio confirms the common-sense judgment
that A2 is far more plausible than A1''
\^ Another posterior argument --- this likelihood ratio in in fact a change
in plausibilities. For example, if A2 is an inconsistent, self-contradictory
theory, it is totally implausible, and not more plausible despite predicting
these data better.
``The statistician can
balance the numerical verdict of likelihood against the vaguer expectations de-
rived from `a priori' considerations, and it must be left to his judgment to
at what point the evidence in favour of alternative hypotheses becomes so con-
vincing that Hypothesis A must be rejected.''
Introduce maximum likelihood ratio as interesting without claiming it to be
necessarily best.
Neyman-Pearson lemma --- reproduce Fig 5?
~\cite{neymanpearson19233lemma}
}

\TODO{
discrete failure to match --- randomization.
table of random numbers --- reputable table!
K. D. Tocher (1950)
``extraneous randomization''
~\cite{tocher1950discontinuous}
}

% TODO Kolmogorov axioms, unbiasedness
\TODO{
Given all this understanding, what I do not understand is their choice to
convert this to a $p$-value.
Seems to be this idea of long-run correctness guarantees.
But their derivation does rely on long-run frequencies.
}

### 2022-08-27 main/draft.tex
add gaussian stuff
add some references for stats section; tweak some data analysis; improve chapter name

>>>>

This is summarized most succinctly by Jaynes:
``any function with a single rounded maximum, raised to a higher and higher
power, goes into a Gaussian function''~\cite{jaynes2003probability}.

More important than this Wald's mathematical proof is that this coincides with
intuition --- when considering decisions, it feels normal to imagine what might
occur in each case and consider how good or bad the plausble outcomes are.

SMRW scan url http://psychclassics.yorku.ca/Fisher/Methods/
